From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-07 Pi+/paper/ufin.tex
Squeeze some space in equations
Change inline sigma notation

  The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
  projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
  fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have

@@

Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
$\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure of
$A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.

@@

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or, $\fst : {\dsum{x:A}{P(x)}} \to A$ is a univalent fibration) if $\tptEqv{P}$ is an
  equivalence.
\end{definition}

@@

\begin{proposition}[Univalent Subuniverse]
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
  for every $X$. The fibration ${\fst : \dsum{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a univalent
  subuniverse
  \[
    \UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)
  \]
\end{proposition}

\begin{appendixproof}
  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{appendixproof}

@@

\begin{definition}
  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is, $\UFin \eqv
    \dsum{n:\Nat}{\UFin[n]}$.
\end{definition}

@@

\begin{proposition}
  For any $n, m : \Nat$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot        \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top        \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]
    \end{aligned}
  \end{gather*}
  and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \bot \sqcup X         & \eqv X                     \\
      X \sqcup \bot         & \eqv X                     \\
      (X \sqcup Y) \sqcup Z & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y            & \eqv Y \sqcup Y
    \end{aligned}
    \qquad
    \begin{aligned}
      \top \times X         & \eqv X                                \\
      X \times \top         & \eqv X                                \\
      (X \times Y) \times Z & \eqv X \times (Y \times Z)            \\
      X \times Y            & \eqv Y \times X                       \\
      X \times \bot         & \eqv \bot                             \\
      X \times (Y \sqcup Z) & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}

>>>>

  The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
  projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
  fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have

@@

Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
$\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure of
$A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.

@@

\begin{definition}[Univalent Fibration]
  $P$ is a univalent type family (or, $\fst : {\dsum*{x:A}{P(x)}} \to A$ is a univalent fibration) if $\tptEqv{P}$ is an
  equivalence.
\end{definition}

@@

\begin{proposition}[Univalent Subuniverse]
  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
  proposition for every $X$. The fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a
  univalent subuniverse
  \[
    \UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)
  \]
\end{proposition}

\begin{appendixproof}
  Suppose $(U, \El) \defeq (\dsum*{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
  inverse follows by calculation using the appropriate computation rules.
\end{appendixproof}

@@

\begin{definition}
  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is,
  $\UFin \eqv \dsum*{n:\Nat}{\UFin[n]}$.
\end{definition}

@@

\begin{proposition}
  For any $n, m : \Nat$,
  \begin{gather*}
    \begin{aligned}
      \Fin[0]                & \eqv \bot        \\
      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]
    \end{aligned}
    \qquad
    \begin{aligned}
      \Fin[1]                & \eqv \top        \\
      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]
    \end{aligned}
  \end{gather*}
  and for any types $X, Y, Z$,
  \begin{gather*}
    \begin{aligned}
      \bot \sqcup X         & \eqv X                     \\
      X \sqcup \bot         & \eqv X                     \\
      (X \sqcup Y) \sqcup Z & \eqv X \sqcup (Y \sqcup Z) \\
      X \sqcup Y            & \eqv Y \sqcup Y            \\
      X \times \bot         & \eqv \bot
    \end{aligned}
    \qquad
    \begin{aligned}
      \top \times X         & \eqv X                                \\
      X \times \top         & \eqv X                                \\
      (X \times Y) \times Z & \eqv X \times (Y \times Z)            \\
      X \times Y            & \eqv Y \times X                       \\
      X \times (Y \sqcup Z) & \eqv (X \times Y) \sqcup (X \times Z)
    \end{aligned}
  \end{gather*}
\end{proposition}
